---
title: "Black Locust Species Distribution Modeling"
author: "Jeffrey Minucci"
date: "7/5/2017"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    fig_height: 5
    fig_width: 7
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig_caption=T)
knitr::opts_chunk$set(message=FALSE, 
tidy.opts=list(width.cutoff=60))

library(dismo)
library(maptools)
library(raster)
library(maps)
library(ggmap)
library(rgdal)
library(caret)
library(SDMTools)
library(gstat)
library(doParallel)
library(pROC)
library(gbm)




#Read data and format
PA <- read.csv("Datafiles/data.national.6.5.csv") #now using reduced factors & elevation
PA$lith.pred <- factor(PA$lith.pred)
PA$Primary.rocktype <- factor(PA$Primary.rocktype)
PA$Secondary.rocktype <- factor(PA$Secondary.rocktype)
PA$usda_tex <- factor(PA$usda_tex)
PA$pres <- factor(PA$pres,levels=c("Present",'Absent'))
PA <- PA[,colnames(PA)!="PlotID"]
PA <- PA[,c(6,9,1:5,7,8,10:length(PA))] #reorder to put responses first
set.seed(1337)
intrain <- createDataPartition(y=PA$pres,p=0.8,list=FALSE)
PA.train.full <- PA[intrain,]
PA.test.full <-  PA[-intrain,]
PA.train <- PA.train.full[,c(-2)]
PA.train <- PA.train[c(2,1:nrow(PA.train)),]  ### Put an absent as the first value
PA.test <- PA.test.full[,c(-2)]
PA.train.c <- PA.train[complete.cases(PA.train),]
sdm1 <- readRDS("Objects/PA_GBM_Models/randBest_6_26_17.rds")
<<<<<<< HEAD

###
# Remove single case where usda_tex = 13 in test set (as there is no usda tex 13 in training set)
#PA.test <- subset(PA.test,usda_tex != "13" | is.na(usda_tex))
#PA.test$usda_tex <- droplevels(PA.test$usda_tex)
=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
```

## Introduction

Symbiotic N~2~-fixing plants can act as biogeochemical keystone species in forest ecosystems, providing massive inputs of nitrogen (N), typically during early succession. As a result, the presence of N~2~-fixers can cause increased N mineralization and nitrification rates, DIN pool sizes in soils, and productivity of non-fixing species (Minucci *et al.*, unpublished).

Despite the importance of symbiotic N~2~-fixers in regulating forest biogeochemistry, we lack an understanding of what factors control the distribution and abundance of these species. As a result, it is difficult to predict how global change factors will alter the presence of N~2~-fixers in forests, potentially leading to declines in some forests and invasion of N~2~-fixers into new regions. 

In the eastern US, one widespread leguminuous N~2~-fixing tree, *R. pseudoacacia* (or black locust) plays a key role in driving forest productivity and recovery from disturbance (Minucci *et al.*, unpublished). Despite its importance, the factors controlling its distribution are poorly understood. The vast majority of leguminous N~2~-fixing tree species are either tropical or subtropical in range and the group is dominant in hot, arid regions. Yet, the distribution of *R. pseudoacacia* is centered in the Appalachian mountains, a wet temperate region. Two previous attempts to model the distribution of *R. pseudoacacia*, either alone (Iversion *et al.*, 2007) or with other leguminous N~2~-fixers had poor performance in predicting current distribution patterns. 


Here we utilized the USDA FS Foresty Inventory and Analysis (FIA) tree demographics dataset and machine learning methods to model *R. pseudoacacia* habitat suitability under current and future climate.

Our goals were to determine :


1. What climate, soil, geological, and forest structure factors drive the distribution of *R. pseudoacacia*?
2. How will the distribution of *R. pseudoacacia* be altered under future climate scenarios?

<<<<<<< HEAD
<br>
<br>

=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
## Methods


### Data sources

<<<<<<< HEAD
<br>
=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

#### Tree demographics: Forest Service FIA data

We used the most recent survey data for naturally regenerated forest plots in the Eastern US. We characterized *R. pseudoacacia* as either present or absent from each plot.

* Total plot number: 82,023 plots 
* Plots with *R. pseudoacacia present: 2586 (3.2%)

<<<<<<< HEAD
```{r echo=FALSE,fig.cap="**All natural forest plots (left), and natural forest plots where *R. pseudoacacia* was present (right)**", fig.width=4, fig.height=4,fig.show='hold'}
usa <- getData('GADM' , country="USA", level=1)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE)
points(LAT~LON,data=PA,pch=3,cex=.01)

=======
```{r echo=FALSE,fig.cap="Naturally regenerated forest plot distribution", fig.width=5, fig.height=5}
usa <- getData('GADM' , country="USA", level=1)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE)
points(LAT~LON,data=PA,pch=3,cex=.01)
```


```{r echo=FALSE,fig.cap="Plots with *R. pseudoacacia* present", fig.width=5, fig.height=5}
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
usa <- getData('GADM' , country="USA", level=1)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE)
points(LAT~LON,data=subset(PA,pres=="Present"),pch=3,cex=.01)
```


<<<<<<< HEAD
<br>


#### Forest structure

From the Forest Service FIA dataset, we extracted data on:

=======
#### Forest structure

From the Forest Service FIA dataset, we extracted data on:
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
* Elevation
* Slope
* Aspect
* Stand Age
* Presence of fire disturbance

We also included annual maximum green vegetation fraction (MGVF), a MODIS-based measurement of greenness (1 km^2^ resolution).  


<<<<<<< HEAD
<br>
=======



>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

#### Climate data

For current climate data, we used **WorldClim 2.0** bioclimatic variables with 30 seconds (~1 km^2^) resolution.

For projected climate in 2050, we used bioclimatic varibles extracted from all **CMIP5** general circulation models under the **RCP 8.5** pathway.

<<<<<<< HEAD
<br>
=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

#### Geological data

To capture differences in parent material, we used:

<<<<<<< HEAD
* **USGS** classifications of parent material primary and secondary rock types

To represent more general differences in surface lithology we used:

* **USGS** maps of surfacial lithology (e.g. clayey glacial till, colluvial sediment)

<br>

#### Soil data

We used the **Harominzed World Soil Database** to extract the following soil features:
=======
* USGS classifications of parent material primary and secondary rock types

To represent more general differences in surface lithology we used:

* USGS maps of surfacial lithology (e.g. clayey glacial till, colluvial sediment)


#### Soil data

We used the Harominzed World Soil Database to extract the following soil features:
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

* pH
* Cation exchange capacity (CEC)
* USDA soil texture classification
* Organic carbon content

<<<<<<< HEAD
<br>
<br>
=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

### Statistical modeling

We used gradient boosted classification trees to model the liklihood of *R. pseudoacacia* presence at each plot. This method allowed us to automate parameter selection and the structure of high-level interactions between parameters. 

<<<<<<< HEAD
<br>

=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
**Potential predictors:**

* BIO1 = Annual Mean Temperature
* BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
* BIO3 = Isothermality (BIO2/BIO7) (* 100)
* BIO4 = Temperature Seasonality (standard deviation *100)
* BIO5 = Max Temperature of Warmest Month
* BIO6 = Min Temperature of Coldest Month
* BIO7 = Temperature Annual Range (BIO5-BIO6)
* BIO8 = Mean Temperature of Wettest Quarter
* BIO9 = Mean Temperature of Driest Quarter
* BIO10 = Mean Temperature of Warmest Quarter
* BIO11 = Mean Temperature of Coldest Quarter
* BIO12 = Annual Precipitation
* BIO13 = Precipitation of Wettest Month
* BIO14 = Precipitation of Driest Month
* BIO15 = Precipitation Seasonality (Coefficient of Variation)
* BIO16 = Precipitation of Wettest Quarter
* BIO17 = Precipitation of Driest Quarter
* BIO18 = Precipitation of Warmest Quarter
* BIO19 = Precipitation of Coldest Quarter
* Longitude
* Latitude
* Elevation
* Stand age
* Slope
* Aspect
* Presence of fire disturbance (yes/no)
* Maximum green vegetation fraction (MGVF)
* Surficial lithography
* Parent material primary rock type
* Parent material secondary rock type
* Soil cation exchange capacity (CEC)
* Soil pH
* Soil USDA texture classification
* Soil organic carbon 

<<<<<<< HEAD
<br>

We split our data into model training (80%) and test (20%) sets. Boosted classification trees were fit to the training set, with optimal hyperparameters determined using five repeats of five-fold cross validation and a random grid search. Goodness of fit was calculated with **AUROC**, the area under the receiver operating characteristic (ROC) curve.

<br>
=======
We split our data into model training (80%) and test (20%) sets. Boosted classification trees were fit to the training set, with optimal hyperparameters determined using five repeats of five-fold cross validation and a random grid search. Goodness of fit was calculated with **AUC**, the area under the receiver operating characteristic (ROC) curve.
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

**Optimial hyperparameters were determined to be:**

* Interaction depth = 32
* Minimum observations in a node = 12
* Number of trees = 800
* Learning rate = 0.01

```{r echo=FALSE,fig.cap="Trajectory of fit versus number of trees", fig.width=5, fig.height=3.5}
ggplot(sdm1,highlight=T,se=T)+theme_bw()+annotate("text",x=830,y=0.914,label="Optimal # of trees")
```

<<<<<<< HEAD
<br>
<br>

## Results

<br>

### Model performance

<br>

Our main assessment of model performance is **AUROC**, or area under the receiver operating characteristic (ROC) curve. This metric takes into account both sensitivity (true positive rate) and specificity (true negative rate) and is preferable to classification accuracy (%) when positive and negative outcomes are imbalanced (and in our case, *R. pseudoacacia* is present in only ~3% of plots).

The best possible AUROC is 1, while the worst possible AUROC (a null model) would be 0.5.

```{r echo=FALSE}
trainROC <- getTrainPerf(sdm1)$TrainROC
predicted.PA.test <- predict(sdm1,newdata=PA.test[,-1],type="prob",na.action=na.pass) 
testROC <- auc(roc(ifelse(PA.test$pres == "Present",1,0),pred=predicted.PA.test$Present))[1]
```

<br>


#### Model accuracy


Metric                               | Value
-----------------------------------  | ----------
Training set AUROC (cross-validated) | `r round(trainROC,3)`
Test set AUROC                       | `r round(testROC,3)`


<br>
<br>



### What factors were most important in determining presence of *R. pseudoacacia*?

<br>

=======
###What factors were most important in determining presence of *R. pseudoacacia*?
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb

Variable                   | Importance 
-------------------------  | ----------
Annual mean temperature    | 100.0
Max temp. of warmest month | 81.2
Primary parent rock type   | 75.1
Slope                      | 41.1
Aspect                     | 34.6
Soil organic carbon        | 33.2

```{r echo=FALSE,fig.cap="Relationship between annual mean temperature and probability of presence", fig.width=5, fig.height=3.5}
plot(plot(sdm1$finalModel,i.var='bio1',type="response",return.grid=T),xlab="Annual mean temp. (C)",
     ylab="Probabiliy of presence",type="l",xaxt="n") 
axis(1,at=c(0,50,100,150,200),labels=c("0","5","10","15","20"))
#rug(PA.train$bio1)
```

<<<<<<< HEAD
```{r echo=FALSE,fig.cap="Relationship between max temp. of warmest month and probability of presence", fig.width=5, fig.height=3.5}
=======
```{r echo=FALSE,fig.cap="Relationship between Max temp. of warming month and probability of presence", fig.width=5, fig.height=3.5}
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
plot(plot(sdm1$finalModel,i.var='bio5',type="response",return.grid=T),xlab="Max temp. of warmest month (C)",
     ylab="Probabiliy of presence",type="l",xaxt="n") #max temp of warmest month
axis(1,at=c(200,250,300,350),labels=c("20","25","30","35"))
#rug(PA.train$bio5)```
```

<<<<<<< HEAD
<br>

#### Parent materials associated with *R. pseudoacacia* presence

```{r echo=FALSE,fig.cap="Relationship between parent material type and probability of presence", fig.width=5, fig.height=3.5}
rockgrid <- plot(sdm1$finalModel,i.var='Primary.rocktype',type="response",return.grid=T)
rockgrid[,1] <- factor(rockgrid[,1],levels=as.character(rockgrid[order(rockgrid[,2]),1]))
plot(rockgrid,ylab="Probability of presence",xlab="Parent material type")
=======
####Parent materials associated with *R. pseudoacacia* presence

```{r echo=FALSE,fig.cap="Relationship between parent material type and probability of presence", fig.width=5, fig.height=3.5}
plot(plot(sdm1$finalModel,i.var='Primary.rocktype',type="response",return.grid=T),ylab="Probability of presence",xlab="Parent material type")
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
#rug(PA.train$bio5)```
```

Parent material type  | Probability of *R. pseudoacacia*
--------------------- | ----------
Biotite schist        | 74.0%
Schist                | 65.1%
Silt                  | 64.1%
All others            | 29.2%



```{r echo=FALSE,fig.cap="Relationship between slope and probability of presence", fig.width=5, fig.height=3.5}
plot(plot(sdm1$finalModel,i.var='SLOPE',type="response",return.grid=T),xlab="Slope (ft rise per 100 ft)",
     ylab="Probabiliy of presence",type="l") #max temp of warmest month
#axis(1,at=c(200,250,300,350),labels=c("20","25","30","35"))
#rug(PA.train$bio5)```
```

```{r echo=FALSE,fig.cap="Relationship between stand age and probability of presence", fig.width=5, fig.height=3.5}
plot(plot(sdm1$finalModel,i.var='STDAGE',type="response",return.grid=T),xlab="Stand age (years)",
     ylab="Probabiliy of presence",type="l") #max temp of warmest month
#axis(1,at=c(200,250,300,350),labels=c("20","25","30","35"))
#rug(PA.train$bio5)
```

```{r echo=FALSE,warning=FALSE,fig.cap="Relationship between soil organic content and probability of presence",fig.width=5, fig.height=3.5}
plot(plot(sdm1$finalModel,i.var='OC',type="response",return.grid=T,continuous.resolution=200),xlim=c(0,2.5),
     ylab="Probabiliy of presence",xlab="Soil organic carbon content (%)",type="l") #max temp of warmest month
#axis(1,at=c(200,250,300,350),labels=c("20","25","30","35"))
rug(PA.train$OC)
```

<<<<<<< HEAD
<br>
<br>

### Using our model to predict current distribution:


```{r, echo=FALSE,warning=FALSE,message=FALSE,fig.width=4.5, fig.height=4,fig.cap="Predicted probability of presence (left) versus actual distribution (right)",fig.show='hold'}
=======

### Using our model to predict current distribution:

```{r, echo=FALSE,warning=FALSE,message=FALSE,fig.width=6, fig.height=4.5,fig.cap="Predicted probability of presence"}
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
# Get probability predictions for ALL DATA (test+train) 
predicted.PA <- predict(sdm1,newdata=PA[,-c(1,2)],type="prob",na.action=na.pass) 
PA.predict <- cbind(PA,predicted.PA)

### Plot probability - as raster
prob.spatial <- PA.predict[,c("LON","LAT","Present")]
names(prob.spatial) <- c("x","y","z")
e <- extent(prob.spatial[,1:2])
r <- raster(e,ncol=130,100)
x <- rasterize(prob.spatial[,1:2],r,prob.spatial[,3],fun=mean)
plot(x,interpolate=T)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE,add=T)
<<<<<<< HEAD


PA$col <- ifelse(PA.predict$pres == "Present","darkgreen","cornflowerblue")
plot(LAT~LON, col=col,data=subset(PA,pres=="Absent"),pch=20,cex=.1)
#xlim=c(-102.5,-62),ylim=c(25,50)
points(LAT~LON, col=col,data=subset(PA,pres!="Absent"),pch=20,cex=.1, asp=1)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE,add=T)
#points(LAT~LON,data=PA,pch=3,cex=.01)

```


<br>
<br>



```{r, echo=FALSE,warning=FALSE,fig.width=6, fig.height=4.5,fig.cap="Plots where *R. pseudoacacia* is present overlaid on predicted probability"}
=======
```

Actual presence/absence:

```{r echo=FALSE,fig.cap="Naturally regenerated forest plot distribution", fig.width=6, fig.height=4.5}
PA$col <- ifelse(PA.predict$pres == "Present","darkgreen","cornflowerblue")
plot(LAT~LON, col=col,data=subset(PA,pres=="Absent"),pch=20,xlim=c(-102.5,-70),ylim=c(25,50),cex=.01)
points(LAT~LON, col=col,data=subset(PA,pres!="Absent"),pch=20,cex=.1)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),add=T)
#points(LAT~LON,data=PA,pch=3,cex=.01)
```


Plots where *R. pseudoacacia* is present overlaid on predited probability:

```{r, echo=FALSE,warning=FALSE,fig.width=6, fig.height=4.5,fig.cap="Predicted probability of presence"}
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
# Get probability predictions for ALL DATA (test+train) 
predicted.PA <- predict(sdm1,newdata=PA[,-c(1,2)],type="prob",na.action=na.pass) 
PA.predict <- cbind(PA,predicted.PA)

### Plot probability - as raster
prob.spatial <- PA.predict[,c("LON","LAT","Present")]
names(prob.spatial) <- c("x","y","z")
e <- extent(prob.spatial[,1:2])
r <- raster(e,ncol=130,100)
x <- rasterize(prob.spatial[,1:2],r,prob.spatial[,3],fun=mean)
plot(x,interpolate=T)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE,add=T)
points(points(LAT~LON,data=subset(PA,pres=="Present"),pch=3,cex=.01))
```

<<<<<<< HEAD
<br>
<br>

### Using our model to predict distribution in 2050

<br>

=======

### Using our model to predict distribution in 2050

>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
Using the average predictions for 17 GCMs (CMIP5 RCP8.5):

```{r, echo=FALSE,warning=FALSE,fig.width=6, fig.height=4.5,fig.cap="Predicted probability of presence in 2050"}
raster_stack <- readRDS("RasterOutput/gcm_predict_rasters_unique.rds")

#Presence/absence probability results (averaged for all GCMs)
gcm_predict_avg <- overlay(raster_stack,fun=mean)
plot(gcm_predict_avg,interpolate=T)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE,add=T)
```

<<<<<<< HEAD
<br>
<br>

=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
Change from present distribution:

```{r, echo=FALSE,warning=FALSE,fig.width=6, fig.height=4.5,fig.cap="Change in probability from current to 2050"}
d_raster_stack <- readRDS("RasterOutput/gcm_predict_rasters_delta_unique.rds")
gcm_d_predict_avg <- overlay(d_raster_stack,fun=mean)
plot(gcm_d_predict_avg,interpolate=T)
plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE,add=T)
```

```{r, echo=FALSE,warning=FALSE,fig.width=6, fig.height=4.5,fig.cap="Change in probability from current to 2050"}
#plot(usa,xlim=c(-90,-70),ylim=(c(25,50)),axes=TRUE)
#points(LAT~LON,data=subset(PA,Primary.rocktype %in% c("72","16","45","15","19","45")),pch=3,cex=.1)
```
<<<<<<< HEAD

<br>
<br>

## Summary of findings

<br>

1. What climate, soil, geological, and forest structure factors drive the distribution of *R. pseudoacacia*?
  + We found that climate, parent material, soil factors, and stand physical characteristics were all important predictors of *R.   pseudoacacia* distribution, with temperature and parent material being the best predictors. Probability of *R. pseudoacacia* occurance was greatest in areas with high mean annual temperature but low maximum summer temperatures and schist or silt geology. *R. pseudoacacia* was also found more frequently in highly sloped areas, potentially due to the greater frequency of disturbance on slopes. 
 
2. How will the distribution of *R. pseudoacacia* be altered under future climate scenarios?
  + Our model predicts a decline in habitat suitability over most of the current range of *R. pseudoacacia*, with the strongest declines in the southern Appalachian region (especially the Ridge and Valley region), the Ozarks, and the Bluegrass region of Kentucky. 
  
  + Our model also predicts a strong increase in habitat suitability at the nothern edge of the range (Michigan, Wisconsin, and central New York), suggesting a potential for *R. pseudoacacia* to expand northward. The invasion of *R. pseudoacacia* could greatly increase N cycling rates and N inputs to these forests. 

=======
>>>>>>> 3d521b03936cc3aa15965652269e102e0117c9cb
